Liveness-Based Garbage Collection for Lazy Languages

### 2015-03-15 paper/paper.tex
pulled parts written by prasanna

A variable is {\em live} it  there is a possibility of its value being
used  in future computation  and dead  if it  is definitely  not used.
Heap-allocated data needs a richer model of liveness which talks about
the liveness  of pointers including  fields of $\CONS$ cells.   We use
$\acar$, $\acdr$ to represent  access using $\CAR$, $\CDR$ fields. The
liveness of the structure reachable  from a variable is represented by
a  set  of  {\em   access  paths}  i.e.   prefix-closed  strings  from
$\{\acar,\acdr\}^\ast$.  As an example, if $x$ is a list with liveness
$\{\epsilon,  \acdr, \acdr\acar, \acdr\acdr,  \acdr\acdr\acar\}$, then
future computations can  only refer up to the  second and third members
of $x$.  The classical notions of a scalar variable being live or dead
correspond to $\{\epsilon\}$ and $\{\}$.

A {\em liveness environment} is a mapping from variables to subsets of
$\{\acar,\acdr\}^\ast$. But  we often express it as a  set, for
example writing $\{x.\acar\acdr, x.\acdr, y.\epsilon\}$ instead of $[x
  \mapsto\{\epsilon,\acar,\acar\acdr,\acdr\},     y\mapsto\{\epsilon\},
  z\mapsto\{\}]$.   For  eager  languages,  we  associate  a  liveness
environment  with each program  point; the  liveness at  program point
$\pi$ in $\pi:e$ is the liveness just before executing $e$. Due to the
inherent uncertainty  in determining the  point of evaluation  in lazy
languages, we  associate liveness  environments with certain  kinds of
paths in the program. All programs  points lying on the path share the
same liveness environment.  We explain this point later.

A complementary  notion to liveness  is {\em demand}.  The  demand for
expression  $e$   is  again  an   access  path---that  subset   of  of
$\{\acar,\acdr\}^\ast$ which  the context of $e$ may  explore of $e$'s
result.  So,  for example given  a demand $\sigma$ and  the expression
$\pi: (\RETURN\ x)$, the liveness at $\pi$ is exactly $x.\sigma$.  The
classical  analogy of  this  is  in {\em  strong  liveness}, where  an
assignment node $n: x:=y+z$ causes $y$  and $z$ to be live on entry to
$n$ if (and only if) $x$ is  live at exit of $n$---the liveness of $x$
at  exit from  $n$ becomes  the demand  on $y+z$.

>>>>

A variable is {\em live} it  there is a possibility of its value being
used  in future computation  and dead  if it  is definitely  not used.
Heap-allocated data needs a richer model of liveness which talks about
the  liveness of pointers  including fields  of $\CONS$  cells.  Using
$\acar$, $\acdr$  to represent access using $\CAR$  and $\CDR$ fields,
the  liveness  of the  structure  reachable  from  a variable  can  be
represented by a set of {\em access paths} i.e.  prefix-closed strings
from $\{\acar,\acdr\}^\ast$.   As an  example, if $x$  is a  list with
liveness      $\{\epsilon,     \acdr,      \acdr\acar,     \acdr\acdr,
\acdr\acdr\acar\}$, then future computations  can only refer up to the
second and third  members of $x$. In this  notation, a scalar variable
being live or dead are represented as $\{\epsilon\}$ and $\{\}$ respectively.

A {\em liveness environment} is a mapping from variables to subsets of
$\{\acar,\acdr\}^\ast$, but  often expressed as a set,  for example by
writing  $\{x.\acar\acdr,   x.\acdr,  y.\epsilon\}$  instead   of  $[x
  \mapsto\{\epsilon,\acar,\acar\acdr,\acdr\},     y\mapsto\{\epsilon\},
  z\mapsto\{\}]$.   We  associate  a  liveness environment  with  each
program  point $\pi$:  the liveness  at $\pi:e$  is the  liveness just
before executing $e$.
%%
%% Due to the inherent uncertainty in
%% determining the  point of evaluation  in lazy languages,  we associate
%% liveness environments with certain kinds  of paths in the program. All
%% programs points lying on the path share the same liveness environment.
%% We explain this point later.
%% 
A  notion  related  to  liveness  is {\em  demand}.   The  demand on
expression  $e$   is  again  an   access  path---that  subset   of  of
$\{\acar,\acdr\}^\ast$ which  the context of $e$ may  explore of $e$'s
result.  So,  for example given  a demand $\sigma$ and  the expression
$\pi: (\RETURN\ x)$, the liveness at $\pi$ is exactly $x.\sigma$.  The
classical  analogy of  this  is  in {\em  strong  liveness}, where  an
assignment node $n: x:=y+z$ causes $y$  and $z$ to be live on entry to
$n$ if (and only if) $x$ is  live at exit of $n$---the liveness of $x$
at exit from $n$ becomes the demand on $y+z$.
